---
abstract: How to update R from one major or minor version to the next (e.g., R 4.4 to R 4.5) by reinstalling packages from the previous version's library, including CRAN packages and packages from GitHub.
execute:
  eval: false
---

# How to update R and reinstall packages

When updating R to a new major or minor version (e.g., from R 4.4 to R 4.5), you will need to reinstall your R packages, as each R version uses a separate library location. This chapter walks you through migrating your packages efficiently.

::: {.callout-important}
## Start with a vanilla session
Before beginning this process, start a clean R session without your usual `.Rprofile` customizations. Follow the instructions in @sec-rprofile to use the vanilla session toggle, or start R with the `--vanilla` flag. This ensures you have a consistent, reproducible session without potential conflicts from your personalized configuration.
:::

## Overview

The process has four main steps:

1. **Identify old packages**: Get a list of packages from your previous R version
2. **Identify GitHub packages**: Automatically detect which packages were installed from GitHub
3. **Install CRAN packages**: Reinstall binary packages from CRAN
4. **Install GitHub packages**: Reinstall packages that came from GitHub

## Step 1: Identify packages from your old R installation

First, we need to identify which packages were in your previous R version. The code below is generic—it will work for any two versions.

```{r}
#| evaluate: false

# Define your old and new R versions
old_version <- "4.4"  # Change to your old version (e.g., "4.3", "4.4")
new_version <- "4.5"  # Change to your new version (e.g., "4.5", "4.6")

# Get the current library path (for the new version)
new_lib <- .libPaths()[1]

# Construct the old library path by replacing version numbers
old_lib <- sub(new_version, old_version, new_lib)

# Verify the old library exists
if (!dir.exists(old_lib)) {
  stop("Old library not found: ", old_lib, 
       "\nCheck that you haven't deleted it and the path is correct.")
}

# Get list of packages from old library
old_packages <- list.files(old_lib)

# Filter out packages already installed in new version
new_packages <- list.files(new_lib)
packages_to_install <- setdiff(old_packages, new_packages)

cat(sprintf("Old R library: %s\n", old_lib))
cat(sprintf("New R library: %s\n", new_lib))
cat(sprintf("Packages to install: %d\n\n", length(packages_to_install)))

# Show the first 10 packages as a preview
if (length(packages_to_install) > 0) {
  cat("First 10 packages to install:\n")
  print(head(packages_to_install, 10))
  if (length(packages_to_install) > 10) {
    cat("... and", length(packages_to_install) - 10, "more\n")
  }
}
```

## Step 2: Identify packages installed from GitHub and separate CRAN packages

Before reinstalling packages, we'll automatically identify which packages in your old library were installed from GitHub. This allows us to use the appropriate installation method for each package type.

```{r}
#| evaluate: false

# Continue from Step 1—old_lib and packages_to_install should be defined

# Function to extract GitHub repository info from old packages
identify_github_packages <- function(lib_path, package_names) {
  
  github_packages <- data.frame(
    package = character(),
    repo = character(),
    stringsAsFactors = FALSE
  )
  
  for (pkg in package_names) {
    pkg_dir <- file.path(lib_path, pkg)
    
    # Check for DESCRIPTION file
    desc_file <- file.path(pkg_dir, "DESCRIPTION")
    if (file.exists(desc_file)) {
      desc <- readLines(desc_file)
      
      # Look for RemoteType and RemoteUrl fields (added by devtools/remotes)
      remote_type <- grep("^RemoteType:", desc, value = TRUE)
      remote_url <- grep("^RemoteUrl:", desc, value = TRUE)
      
      if (length(remote_type) > 0 && length(remote_url) > 0) {
        # Extract repository spec from URL
        # GitHub URLs typically end with owner/repo.git
        if (grepl("github.com", remote_url)) {
          # Extract "owner/repo" from URL like https://github.com/owner/repo.git
          repo_match <- gsub(".*github.com/(.+?)(\\.git)?/?$", "\\1", remote_url)
          
          # Verify it looks like owner/repo format
          if (grepl("^[^/]+/[^/]+$", repo_match)) {
            github_packages <- rbind(
              github_packages,
              data.frame(package = pkg, repo = repo_match, stringsAsFactors = FALSE)
            )
          }
        }
      }
    }
  }
  
  github_packages
}

# Identify GitHub packages in old library
cat("Identifying packages installed from GitHub...\n\n")
github_packages_df <- identify_github_packages(old_lib, packages_to_install)

if (nrow(github_packages_df) > 0) {
  cat(sprintf("Found %d packages from GitHub:\n\n", nrow(github_packages_df)))
  print(github_packages_df)
  cat("\n")
  
  # Save as a vector for use in Step 3
  github_packages_vector <- github_packages_df$repo
} else {
  cat("No GitHub packages found in old library.\n\n")
  github_packages_vector <- character(0)
}

# Remove GitHub packages from CRAN packages list
# (they'll be handled separately in Step 3)
packages_to_install_cran <- setdiff(
  packages_to_install,
  github_packages_df$package
)

cat(sprintf("CRAN packages to install: %d\n", length(packages_to_install_cran)))
cat(sprintf("GitHub packages to install: %d\n\n", length(github_packages_vector)))
```

## Step 3: Install CRAN packages

Once you've separated the packages, install the CRAN packages. The script below handles failures gracefully, skipping packages that can't be installed and logging the results.

```{r}
#| evaluate: false

# Continue from Step 2—packages_to_install_cran should be defined

# Set options to avoid interactive prompts
options(
  install.packages.check.source = "no",
  Ncpus = parallel::detectCores(logical = TRUE)
)
Sys.setenv(R_INSTALL_PKG_EDIT_MODE = "never")

# Track installation results
results <- data.frame(
  package = packages_to_install_cran,
  status = NA_character_,
  stringsAsFactors = FALSE
)

# Install each package, skipping failures
for (i in seq_along(packages_to_install_cran)) {
  pkg <- packages_to_install_cran[i]
  cat(sprintf("[%d/%d] Installing %s... ", i, length(packages_to_install_cran), pkg))
  
  result <- tryCatch(
    {
      install.packages(
        pkg,
        repos = "https://cran.rstudio.com/",
        type = "binary",
        dependencies = FALSE,
        verbose = FALSE,
        quiet = TRUE
      )
      "✓ Success"
    },
    error = function(e) {
      paste("✗ Failed:", e$message)
    },
    warning = function(w) {
      "⚠ Warning"
    }
  )
  
  results$status[i] <- result
  cat(result, "\n")
}

# Summary
cat("\n")
cat("=== Installation Summary ===\n")
table(results$status)

# Show failed packages for manual review
failed <- results$package[grepl("Failed", results$status)]
if (length(failed) > 0) {
  cat("\nFailed packages (may need manual installation or GitHub source):\n")
  print(failed)
}
```

## Step 4: Install packages from GitHub

Packages identified as being from GitHub in Step 2 will be installed here automatically. If you have additional GitHub packages to install, you can specify them manually.

::: {.callout-note}
## Why not use `pak`?
The `pak` package provides excellent package management features, but some institutions restrict its use for security reasons. The methods below use base R and `devtools`, which are more widely available.
:::

### Automatic installation (from Step 2)

If you completed Step 2, your GitHub packages are automatically identified and can be installed here:

```{r}
#| evaluate: false

# Continue from Step 2—github_packages_vector should be defined
# (or it will be an empty vector if no GitHub packages were found)

# Install devtools if not already available
if (length(github_packages_vector) > 0 && 
    !require("devtools", quietly = TRUE)) {
  cat("Installing devtools for GitHub package support...\n")
  install.packages("devtools", repos = "https://cran.rstudio.com/")
}

if (length(github_packages_vector) > 0) {
  cat(sprintf("\nInstalling %d packages from GitHub...\n\n", 
              length(github_packages_vector)))
  
  github_results <- data.frame(
    repo = github_packages_vector,
    status = NA_character_,
    stringsAsFactors = FALSE
  )
  
  for (i in seq_along(github_packages_vector)) {
    repo_spec <- github_packages_vector[i]
    cat(sprintf("[%d/%d] Installing %s... ", 
                i, length(github_packages_vector), repo_spec))
    
    result <- tryCatch(
      {
        devtools::install_github(
          repo_spec,
          dependencies = FALSE,
          type = "binary",
          quiet = TRUE
        )
        "✓"
      },
      error = function(e) "✗"
    )
    
    github_results$status[i] <- result
    cat(result, "\n")
  }
  
  cat("\n=== GitHub Installation Summary ===\n")
  print(table(github_results$status))
  
} else {
  cat("No GitHub packages to install.\n")
}
```

### Manual installation (for additional packages)

If you have additional GitHub packages not identified in Step 2, or if automatic installation fails, you can install them manually using one of these methods.

#### Method 1: Using `devtools::install_github()` (Recommended)

This is the most straightforward method if you have `devtools` installed.

```{r}
#| evaluate: false

# Install devtools if not already available
if (!require("devtools", quietly = TRUE)) {
  install.packages("devtools", repos = "https://cran.rstudio.com/")
}

# Manually specify additional GitHub packages
# Format: "owner/repo" or "owner/repo@branch"
manual_github_packages <- c(
  # "tidyverse/ggplot2",           # Latest development version
  # "r-lib/devtools",              # From r-lib organization
  # "hadley/ggplot2@main",         # Specific branch
  # "owner/private-repo",          # Private repos (with credentials)
)

if (length(manual_github_packages) > 0) {
  cat(sprintf("Installing %d manually-specified packages from GitHub...\n\n",
              length(manual_github_packages)))
  
  for (i in seq_along(manual_github_packages)) {
    pkg_spec <- manual_github_packages[i]
    cat(sprintf("[%d/%d] Installing %s... ",
                i, length(manual_github_packages), pkg_spec))
    
    result <- tryCatch(
      {
        devtools::install_github(
          pkg_spec,
          dependencies = FALSE,
          type = "binary",
          quiet = TRUE
        )
        "✓"
      },
      error = function(e) "✗"
    )
    
    cat(result, "\n")
  }
} else {
  cat("No manual GitHub packages specified.\n")
}
```

### Method 2: Direct download and installation

If `devtools` is unavailable or GitHub packages have dependencies, you can download and install directly:

```{r}
#| evaluate: false

# Create a temporary directory for downloads
temp_dir <- file.path(tempdir(), "gh_packages")
dir.create(temp_dir, showWarnings = FALSE)

# Specify GitHub packages as "owner/repo"
github_packages <- c(
  # "tidyverse/ggplot2",
  # "r-lib/devtools"
)

if (length(github_packages) > 0) {
  for (pkg_spec in github_packages) {
    parts <- strsplit(pkg_spec, "@")[[1]]
    repo_path <- parts[1]  # "owner/repo"
    branch <- if (length(parts) > 1) parts[2] else "main"
    
    # Construct download URL
    url <- sprintf("https://github.com/%s/archive/refs/heads/%s.zip",
                   repo_path, branch)
    
    pkg_name <- basename(strsplit(repo_path, "/")[[1]][2])
    zip_file <- file.path(temp_dir, paste0(pkg_name, ".zip"))
    
    cat(sprintf("Downloading %s from %s...\n", pkg_name, url))
    
    tryCatch(
      {
        download.file(url, zip_file, quiet = TRUE, mode = "wb")
        
        # Extract and install
        extract_dir <- file.path(temp_dir, pkg_name)
        unzip(zip_file, exdir = extract_dir)
        
        # Find the package directory (should be owner-repo-branch/)
        pkg_source <- list.files(extract_dir, full.names = TRUE)[1]
        
        cat(sprintf("Installing %s...\n", pkg_name))
        install.packages(
          pkg_source,
          repos = NULL,
          type = "source",
          dependencies = FALSE
        )
        cat("✓ Successfully installed\n")
      },
      error = function(e) {
        cat("✗ Failed:", e$message, "\n")
      }
    )
  }
  
  # Cleanup
  unlink(temp_dir, recursive = TRUE)
}
```

### Method 3: From source using R CMD INSTALL

For maximum compatibility, you can use R's built-in source installation:

```{r}
#| evaluate: false

# Function to install GitHub package from source
install_github_source <- function(repo, branch = "main", 
                                 install_dir = tempdir()) {
  
  # Parse repo spec
  parts <- strsplit(repo, "@")[[1]]
  repo_path <- parts[1]
  if (length(parts) > 1) branch <- parts[2]
  
  pkg_name <- basename(strsplit(repo_path, "/")[[1]][2])
  
  # Download URL
  url <- sprintf("https://github.com/%s/archive/refs/heads/%s.tar.gz",
                 repo_path, branch)
  
  tar_file <- file.path(install_dir, paste0(pkg_name, ".tar.gz"))
  pkg_dir <- file.path(install_dir, pkg_name)
  
  cat(sprintf("Downloading %s...\n", pkg_name))
  
  tryCatch(
    {
      download.file(url, tar_file, quiet = TRUE)
      
      # Extract
      untar(tar_file, exdir = install_dir)
      
      # Find actual extracted directory (has -branch suffix)
      extracted <- list.files(install_dir, 
                             pattern = paste0(pkg_name, "-"),
                             full.names = TRUE)[1]
      
      cat(sprintf("Installing %s from source...\n", pkg_name))
      
      # Use shell to run R CMD INSTALL
      system(sprintf("R CMD INSTALL \"%s\"", extracted))
      
      cat("✓ Installation complete\n")
    },
    error = function(e) {
      cat("✗ Failed:", e$message, "\n")
    }
  )
}

# Example usage:
# install_github_source("tidyverse/ggplot2")
# install_github_source("r-lib/devtools@main")
```

## Complete workflow script

Here's a complete, integrated script that combines all steps with automatic GitHub package detection:

```{r}
#| evaluate: false
#| code-fold: true

# ==============================================================================
# R Package Migration Script (with Automatic GitHub Detection)
# ==============================================================================
# Migrate packages from an old R version to a new one, automatically
# identifying and installing GitHub packages
# ==============================================================================

# Configuration
OLD_VERSION <- "4.4"   # Your old R version
NEW_VERSION <- "4.5"   # Your new R version

# Optional: Additional GitHub packages to install manually
# Format: "owner/repo" or "owner/repo@branch"
ADDITIONAL_GITHUB_PACKAGES <- c(
  # "tidyverse/ggplot2",
  # "r-lib/devtools@main",
  NULL
)

# ==============================================================================
# STEP 1: IDENTIFY PACKAGES
# ==============================================================================

cat("Step 1: Identifying packages from old R installation\n")
cat("=".^50, "\n\n")

new_lib <- .libPaths()[1]
old_lib <- sub(NEW_VERSION, OLD_VERSION, new_lib)

if (!dir.exists(old_lib)) {
  stop("Old library not found: ", old_lib)
}

old_packages <- list.files(old_lib)
new_packages <- list.files(new_lib)
packages_to_install <- setdiff(old_packages, new_packages)

cat(sprintf("Old R version library: %s\n", old_lib))
cat(sprintf("New R version library: %s\n", new_lib))
cat(sprintf("Total packages to install: %d\n\n", length(packages_to_install)))

if (length(packages_to_install) == 0) {
  cat("No packages to install. All packages already in new version.\n")
  q("no")
}

# ==============================================================================
# STEP 2b: IDENTIFY GITHUB PACKAGES AUTOMATICALLY
# ==============================================================================

cat("Step 2b: Identifying packages from GitHub...\n")
cat("=".^50, "\n\n")

# Function to extract GitHub repository info from package DESCRIPTION
identify_github_packages <- function(lib_path, package_names) {
  github_packages <- data.frame(
    package = character(),
    repo = character(),
    stringsAsFactors = FALSE
  )
  
  for (pkg in package_names) {
    pkg_dir <- file.path(lib_path, pkg)
    desc_file <- file.path(pkg_dir, "DESCRIPTION")
    
    if (file.exists(desc_file)) {
      desc <- readLines(desc_file)
      
      # Look for RemoteType and RemoteUrl fields
      remote_type <- grep("^RemoteType:", desc, value = TRUE)
      remote_url <- grep("^RemoteUrl:", desc, value = TRUE)
      
      if (length(remote_type) > 0 && length(remote_url) > 0) {
        if (grepl("github.com", remote_url)) {
          # Extract "owner/repo" from URL
          repo_match <- gsub(".*github.com/(.+?)(\\.git)?/?$", "\\1", remote_url)
          
          if (grepl("^[^/]+/[^/]+$", repo_match)) {
            github_packages <- rbind(
              github_packages,
              data.frame(package = pkg, repo = repo_match, 
                        stringsAsFactors = FALSE)
            )
          }
        }
      }
    }
  }
  
  github_packages
}

# Identify GitHub packages
github_packages_df <- identify_github_packages(old_lib, packages_to_install)

if (nrow(github_packages_df) > 0) {
  cat(sprintf("Found %d packages from GitHub:\n\n", nrow(github_packages_df)))
  print(github_packages_df)
  cat("\n")
  github_packages_auto <- github_packages_df$repo
} else {
  cat("No GitHub packages found in old library.\n\n")
  github_packages_auto <- character(0)
}

# Separate CRAN and GitHub packages
packages_to_install_cran <- setdiff(
  packages_to_install,
  github_packages_df$package
)

cat(sprintf("CRAN packages to install: %d\n", length(packages_to_install_cran)))
cat(sprintf("GitHub packages to install: %d\n\n", length(github_packages_auto)))

# ==============================================================================
# STEP 2: INSTALL CRAN PACKAGES
# ==============================================================================

cat("Step 2: Installing CRAN packages\n")
cat("=".^50, "\n\n")

options(
  install.packages.check.source = "no",
  Ncpus = parallel::detectCores(logical = TRUE)
)
Sys.setenv(R_INSTALL_PKG_EDIT_MODE = "never")

cran_results <- data.frame(
  package = packages_to_install_cran,
  status = NA_character_,
  stringsAsFactors = FALSE
)

if (nrow(cran_results) > 0) {
  for (i in seq_along(packages_to_install_cran)) {
    pkg <- packages_to_install_cran[i]
    cat(sprintf("[%d/%d] %s... ", i, length(packages_to_install_cran), pkg))
    
    result <- tryCatch(
      {
        install.packages(
          pkg,
          repos = "https://cran.rstudio.com/",
          type = "binary",
          dependencies = FALSE,
          verbose = FALSE,
          quiet = TRUE
        )
        "✓"
      },
      error = function(e) "✗",
      warning = function(w) "⚠"
    )
    
    cran_results$status[i] <- result
    cat(result, "\n")
  }
  
  # Summary
  cat("\n=== CRAN Installation Summary ===\n")
  print(table(cran_results$status))
  
  cran_failed <- cran_results$package[cran_results$status == "✗"]
  if (length(cran_failed) > 0) {
    cat("\nFailed CRAN packages:\n")
    print(cran_failed)
  }
} else {
  cat("No CRAN packages to install.\n")
}

# ==============================================================================
# STEP 3: INSTALL GITHUB PACKAGES (Automatic + Manual)
# ==============================================================================

# Combine automatic and manual GitHub packages
ADDITIONAL_GITHUB_PACKAGES <- Filter(Negate(is.null), ADDITIONAL_GITHUB_PACKAGES)
all_github_packages <- c(github_packages_auto, ADDITIONAL_GITHUB_PACKAGES)
all_github_packages <- unique(all_github_packages)

if (length(all_github_packages) > 0) {
  cat("\n\nStep 3: Installing GitHub packages\n")
  cat("=".^50, "\n\n")
  
  # Install devtools if needed
  if (!require("devtools", quietly = TRUE)) {
    cat("Installing devtools for GitHub package support...\n")
    install.packages("devtools", repos = "https://cran.rstudio.com/")
  }
  
  cat(sprintf("Installing %d packages from GitHub...\n\n", 
              length(all_github_packages)))
  
  github_results <- data.frame(
    repo = all_github_packages,
    status = NA_character_,
    stringsAsFactors = FALSE
  )
  
  for (i in seq_along(all_github_packages)) {
    repo_spec <- all_github_packages[i]
    cat(sprintf("[%d/%d] %s... ", i, length(all_github_packages), repo_spec))
    
    result <- tryCatch(
      {
        devtools::install_github(
          repo_spec,
          dependencies = FALSE,
          type = "binary",
          quiet = TRUE
        )
        "✓"
      },
      error = function(e) "✗"
    )
    
    github_results$status[i] <- result
    cat(result, "\n")
  }
  
  cat("\n=== GitHub Installation Summary ===\n")
  print(table(github_results$status))
  
  github_failed <- github_results$repo[github_results$status == "✗"]
  if (length(github_failed) > 0) {
    cat("\nFailed GitHub packages:\n")
    print(github_failed)
  }
} else {
  cat("\n\nStep 3: No GitHub packages to install.\n")
}

# ==============================================================================
# COMPLETION
# ==============================================================================

cat("\n\n")
cat("=".^50, "\n")
cat("Package migration complete!\n")
cat("=".^50, "\n")

n_cran <- if (nrow(cran_results) > 0) {
  sum(cran_results$status %in% c("✓", "⚠"))
} else {
  0
}

n_github <- if (exists("github_results") && length(all_github_packages) > 0) {
  sum(github_results$status == "✓")
} else {
  0
}

cat(sprintf("Installed %d packages from CRAN\n", n_cran))
cat(sprintf("Installed %d packages from GitHub\n", n_github))
cat("Review any failed packages above and install manually if needed.\n")
```

## Important notes

### Before you start

1. **Back up your old library**: Don't delete your old R library until you've verified everything works
2. **Start with vanilla session**: Use the vanilla session toggle from @sec-rprofile to avoid conflicts
3. **Check internet connection**: Package installation requires stable internet
4. **Have admin rights**: You may need administrator access to write to library directories

### Troubleshooting

**"Old library not found"**
: Make sure you haven't deleted the old R library, and check that you've set the version numbers correctly.

**Binary packages not available**
: Some packages may only be available as source. Add `type = "source"` to `install.packages()`, but note that this requires build tools (Rtools on Windows).

**GitHub authentication errors**
: If you're installing private repositories, make sure you have GitHub credentials set up. See `?devtools::install_github` for details.

**Package conflicts**
: If packages fail to install, check for missing dependencies. You can install them manually first with `install.packages("dependency_name")`.

### After installation

- Test your packages by loading a few: `library(package_name)`
- Run any diagnostic scripts that depend on your packages
- Keep the old library for a few days before deletion, in case you need to reference it
- Update your package migration script for next time (add any GitHub packages you installed manually)

## Next steps

Once your packages are installed, remember to:

1. Return to regular session mode (vanilla mode automatically toggles back)
2. Review @sec-rprofile to ensure your startup configuration is optimal
3. Consider using `renv` for project-specific package management if reproducibility is critical

