---
abstract: How to update R from one major or minor version to the next (e.g., R 4.4 to R 4.5) by reinstalling packages from the previous version's library, including CRAN packages and packages from GitHub.
execute:
  eval: false
---

# How to update R and reinstall packages

When updating R to a new major or minor version (e.g., from R 4.4 to R 4.5), you will need to reinstall your R packages, as each R version uses a separate library location. This chapter walks you through migrating your packages efficiently.

::: {.callout-important}
## Start with a vanilla session
Before beginning this process, start a clean R session without your usual `.Rprofile` customizations. Follow the instructions in @sec-rprofile to use the vanilla session toggle, or start R with the `--vanilla` flag. This ensures you have a consistent, reproducible session without potential conflicts from your personalized configuration.
:::

## Overview

The process has three main steps:

1. **Identify old packages**: Get a list of packages from your previous R version
2. **Install CRAN packages**: Reinstall binary packages from CRAN
3. **Install GitHub packages**: Handle packages that need to be installed from GitHub

## Step 1: Identify packages from your old R installation

First, we need to identify which packages were in your previous R version. The code below is generic—it will work for any two versions.

```{r}
#| evaluate: false

# Define your old and new R versions
old_version <- "4.4"  # Change to your old version (e.g., "4.3", "4.4")
new_version <- "4.5"  # Change to your new version (e.g., "4.5", "4.6")

# Get the current library path (for the new version)
new_lib <- .libPaths()[1]

# Construct the old library path by replacing version numbers
old_lib <- sub(new_version, old_version, new_lib)

# Verify the old library exists
if (!dir.exists(old_lib)) {
  stop("Old library not found: ", old_lib, 
       "\nCheck that you haven't deleted it and the path is correct.")
}

# Get list of packages from old library
old_packages <- list.files(old_lib)

# Filter out packages already installed in new version
new_packages <- list.files(new_lib)
packages_to_install <- setdiff(old_packages, new_packages)

cat(sprintf("Old R library: %s\n", old_lib))
cat(sprintf("New R library: %s\n", new_lib))
cat(sprintf("Packages to install: %d\n\n", length(packages_to_install)))

# Show the first 10 packages as a preview
if (length(packages_to_install) > 0) {
  cat("First 10 packages to install:\n")
  print(head(packages_to_install, 10))
  if (length(packages_to_install) > 10) {
    cat("... and", length(packages_to_install) - 10, "more\n")
  }
}
```

## Step 2: Install CRAN packages

Once you've identified the packages, install them from CRAN. The script below handles failures gracefully, skipping packages that can't be installed and logging the results.

```{r}
#| evaluate: false

# Continue from Step 1—packages_to_install should be defined

# Set options to avoid interactive prompts
options(
  install.packages.check.source = "no",
  Ncpus = parallel::detectCores(logical = TRUE)
)
Sys.setenv(R_INSTALL_PKG_EDIT_MODE = "never")

# Track installation results
results <- data.frame(
  package = packages_to_install,
  status = NA_character_,
  stringsAsFactors = FALSE
)

# Install each package, skipping failures
for (i in seq_along(packages_to_install)) {
  pkg <- packages_to_install[i]
  cat(sprintf("[%d/%d] Installing %s... ", i, length(packages_to_install), pkg))
  
  result <- tryCatch(
    {
      install.packages(
        pkg,
        repos = "https://cran.rstudio.com/",
        type = "binary",
        dependencies = FALSE,
        verbose = FALSE,
        quiet = TRUE
      )
      "✓ Success"
    },
    error = function(e) {
      paste("✗ Failed:", e$message)
    },
    warning = function(w) {
      "⚠ Warning"
    }
  )
  
  results$status[i] <- result
  cat(result, "\n")
}

# Summary
cat("\n")
cat("=== Installation Summary ===\n")
table(results$status)

# Show failed packages for manual review
failed <- results$package[grepl("Failed", results$status)]
if (length(failed) > 0) {
  cat("\nFailed packages (may need manual installation or GitHub source):\n")
  print(failed)
}
```

## Step 3: Install packages from GitHub

Some packages may not be available on CRAN or may require the development version from GitHub. We'll install these using base R tools (without `pak`, which may not be available in your institution).

::: {.callout-note}
## Why not use `pak`?
The `pak` package provides excellent package management features, but some institutions restrict its use for security reasons. The methods below use base R and `devtools`, which are more widely available.
:::

### Method 1: Using `devtools::install_github()` (Recommended)

This is the most straightforward method if you have `devtools` installed.

```{r}
#| evaluate: false

# Install devtools if not already available
if (!require("devtools", quietly = TRUE)) {
  install.packages("devtools", repos = "https://cran.rstudio.com/")
}

# List of packages to install from GitHub
# Format: "owner/repo" or "owner/repo@branch"
github_packages <- list(
  # Example entries—replace with your actual packages
  "tidyverse/ggplot2",           # Latest development version
  "r-lib/devtools",              # From r-lib organization
  # "hadley/ggplot2@main",       # Specific branch
  # "owner/private-repo",        # Private repos work if you have credentials
  NULL  # Placeholder to make it easy to add more
)

# Remove NULL entries and duplicates
github_packages <- Filter(Negate(is.null), github_packages)

if (length(github_packages) > 0) {
  # Install each GitHub package
  for (i in seq_along(github_packages)) {
    pkg_spec <- github_packages[[i]]
    cat(sprintf("\n[%d/%d] Installing from GitHub: %s\n", 
                i, length(github_packages), pkg_spec))
    
    tryCatch(
      {
        devtools::install_github(
          pkg_spec,
          dependencies = FALSE,
          type = "binary",
          quiet = TRUE
        )
        cat("✓ Successfully installed\n")
      },
      error = function(e) {
        cat("✗ Failed to install:", e$message, "\n")
      }
    )
  }
} else {
  cat("No GitHub packages specified.\n")
}
```

### Method 2: Direct download and installation

If `devtools` is unavailable or GitHub packages have dependencies, you can download and install directly:

```{r}
#| evaluate: false

# Create a temporary directory for downloads
temp_dir <- file.path(tempdir(), "gh_packages")
dir.create(temp_dir, showWarnings = FALSE)

# Specify GitHub packages as "owner/repo"
github_packages <- c(
  # "tidyverse/ggplot2",
  # "r-lib/devtools"
)

if (length(github_packages) > 0) {
  for (pkg_spec in github_packages) {
    parts <- strsplit(pkg_spec, "@")[[1]]
    repo_path <- parts[1]  # "owner/repo"
    branch <- if (length(parts) > 1) parts[2] else "main"
    
    # Construct download URL
    url <- sprintf("https://github.com/%s/archive/refs/heads/%s.zip",
                   repo_path, branch)
    
    pkg_name <- basename(strsplit(repo_path, "/")[[1]][2])
    zip_file <- file.path(temp_dir, paste0(pkg_name, ".zip"))
    
    cat(sprintf("Downloading %s from %s...\n", pkg_name, url))
    
    tryCatch(
      {
        download.file(url, zip_file, quiet = TRUE, mode = "wb")
        
        # Extract and install
        extract_dir <- file.path(temp_dir, pkg_name)
        unzip(zip_file, exdir = extract_dir)
        
        # Find the package directory (should be owner-repo-branch/)
        pkg_source <- list.files(extract_dir, full.names = TRUE)[1]
        
        cat(sprintf("Installing %s...\n", pkg_name))
        install.packages(
          pkg_source,
          repos = NULL,
          type = "source",
          dependencies = FALSE
        )
        cat("✓ Successfully installed\n")
      },
      error = function(e) {
        cat("✗ Failed:", e$message, "\n")
      }
    )
  }
  
  # Cleanup
  unlink(temp_dir, recursive = TRUE)
}
```

### Method 3: From source using R CMD INSTALL

For maximum compatibility, you can use R's built-in source installation:

```{r}
#| evaluate: false

# Function to install GitHub package from source
install_github_source <- function(repo, branch = "main", 
                                 install_dir = tempdir()) {
  
  # Parse repo spec
  parts <- strsplit(repo, "@")[[1]]
  repo_path <- parts[1]
  if (length(parts) > 1) branch <- parts[2]
  
  pkg_name <- basename(strsplit(repo_path, "/")[[1]][2])
  
  # Download URL
  url <- sprintf("https://github.com/%s/archive/refs/heads/%s.tar.gz",
                 repo_path, branch)
  
  tar_file <- file.path(install_dir, paste0(pkg_name, ".tar.gz"))
  pkg_dir <- file.path(install_dir, pkg_name)
  
  cat(sprintf("Downloading %s...\n", pkg_name))
  
  tryCatch(
    {
      download.file(url, tar_file, quiet = TRUE)
      
      # Extract
      untar(tar_file, exdir = install_dir)
      
      # Find actual extracted directory (has -branch suffix)
      extracted <- list.files(install_dir, 
                             pattern = paste0(pkg_name, "-"),
                             full.names = TRUE)[1]
      
      cat(sprintf("Installing %s from source...\n", pkg_name))
      
      # Use shell to run R CMD INSTALL
      system(sprintf("R CMD INSTALL \"%s\"", extracted))
      
      cat("✓ Installation complete\n")
    },
    error = function(e) {
      cat("✗ Failed:", e$message, "\n")
    }
  )
}

# Example usage:
# install_github_source("tidyverse/ggplot2")
# install_github_source("r-lib/devtools@main")
```

## Complete workflow script

Here's a complete, integrated script that combines all steps:

```{r}
#| evaluate: false
#| code-fold: true

# ==============================================================================
# R Package Migration Script
# ==============================================================================
# Migrate packages from an old R version to a new one
# ==============================================================================

# Configuration
OLD_VERSION <- "4.4"   # Your old R version
NEW_VERSION <- "4.5"   # Your new R version

# GitHub packages to install (if any)
GITHUB_PACKAGES <- c(
  # "tidyverse/ggplot2",
  # "r-lib/devtools",
  NULL
)

# ==============================================================================
# STEP 1: IDENTIFY PACKAGES
# ==============================================================================

cat("Step 1: Identifying packages from old R installation\n")
cat("=".^50, "\n\n")

new_lib <- .libPaths()[1]
old_lib <- sub(NEW_VERSION, OLD_VERSION, new_lib)

if (!dir.exists(old_lib)) {
  stop("Old library not found: ", old_lib)
}

old_packages <- list.files(old_lib)
new_packages <- list.files(new_lib)
packages_to_install <- setdiff(old_packages, new_packages)

cat(sprintf("Old R version library: %s\n", old_lib))
cat(sprintf("New R version library: %s\n", new_lib))
cat(sprintf("Total packages to install: %d\n\n", length(packages_to_install)))

if (length(packages_to_install) == 0) {
  cat("No packages to install. All packages already in new version.\n")
  q("no")
}

# ==============================================================================
# STEP 2: INSTALL CRAN PACKAGES
# ==============================================================================

cat("Step 2: Installing CRAN packages\n")
cat("=".^50, "\n\n")

options(
  install.packages.check.source = "no",
  Ncpus = parallel::detectCores(logical = TRUE)
)
Sys.setenv(R_INSTALL_PKG_EDIT_MODE = "never")

results <- data.frame(
  package = packages_to_install,
  status = NA_character_,
  stringsAsFactors = FALSE
)

for (i in seq_along(packages_to_install)) {
  pkg <- packages_to_install[i]
  cat(sprintf("[%d/%d] %s... ", i, length(packages_to_install), pkg))
  
  result <- tryCatch(
    {
      install.packages(
        pkg,
        repos = "https://cran.rstudio.com/",
        type = "binary",
        dependencies = FALSE,
        verbose = FALSE,
        quiet = TRUE
      )
      "✓"
    },
    error = function(e) "✗",
    warning = function(w) "⚠"
  )
  
  results$status[i] <- result
  cat(result, "\n")
}

# Summary
cat("\n=== CRAN Installation Summary ===\n")
print(table(results$status))

failed <- results$package[results$status == "✗"]
if (length(failed) > 0) {
  cat("\nFailed packages:\n")
  print(failed)
}

# ==============================================================================
# STEP 3: INSTALL GITHUB PACKAGES (if specified)
# ==============================================================================

GITHUB_PACKAGES <- Filter(Negate(is.null), GITHUB_PACKAGES)

if (length(GITHUB_PACKAGES) > 0) {
  cat("\n\nStep 3: Installing GitHub packages\n")
  cat("=".^50, "\n\n")
  
  if (!require("devtools", quietly = TRUE)) {
    cat("Installing devtools for GitHub package support...\n")
    install.packages("devtools", repos = "https://cran.rstudio.com/")
  }
  
  for (i in seq_along(GITHUB_PACKAGES)) {
    pkg_spec <- GITHUB_PACKAGES[i]
    cat(sprintf("[%d/%d] %s... ", i, length(GITHUB_PACKAGES), pkg_spec))
    
    result <- tryCatch(
      {
        devtools::install_github(
          pkg_spec,
          dependencies = FALSE,
          type = "binary",
          quiet = TRUE
        )
        "✓"
      },
      error = function(e) "✗"
    )
    
    cat(result, "\n")
  }
}

# ==============================================================================
# COMPLETION
# ==============================================================================

cat("\n\n")
cat("=".^50, "\n")
cat("Package migration complete!\n")
cat("=".^50, "\n")
cat(sprintf("Installed %d packages from CRAN\n", 
            sum(results$status %in% c("✓", "⚠"))))
cat("Review any failed packages above and install manually if needed.\n")
```

## Important notes

### Before you start

1. **Back up your old library**: Don't delete your old R library until you've verified everything works
2. **Start with vanilla session**: Use the vanilla session toggle from @sec-rprofile to avoid conflicts
3. **Check internet connection**: Package installation requires stable internet
4. **Have admin rights**: You may need administrator access to write to library directories

### Troubleshooting

**"Old library not found"**
: Make sure you haven't deleted the old R library, and check that you've set the version numbers correctly.

**Binary packages not available**
: Some packages may only be available as source. Add `type = "source"` to `install.packages()`, but note that this requires build tools (Rtools on Windows).

**GitHub authentication errors**
: If you're installing private repositories, make sure you have GitHub credentials set up. See `?devtools::install_github` for details.

**Package conflicts**
: If packages fail to install, check for missing dependencies. You can install them manually first with `install.packages("dependency_name")`.

### After installation

- Test your packages by loading a few: `library(package_name)`
- Run any diagnostic scripts that depend on your packages
- Keep the old library for a few days before deletion, in case you need to reference it
- Update your package migration script for next time (add any GitHub packages you installed manually)

## Next steps

Once your packages are installed, remember to:

1. Return to regular session mode (vanilla mode automatically toggles back)
2. Review @sec-rprofile to ensure your startup configuration is optimal
3. Consider using `renv` for project-specific package management if reproducibility is critical